Synthesizing Quotient Lenses

### 2018-02-23 papers/QuotientLenses/main.tex
minor edits

\emph{Lenses} define computations that can be read in two directions, forwards and
backwards.  The simplest lenses define bijections between source and
data, but we are interested in \emph{quotient lenses}, which define
bijections modulo some equivalence relation.  The remains of this section
describe and motivate a new language and system for defining such lenses.

@@

While a projection chooses a single canonical representative for all elements
of a regular language, the \kw{squash} combinator maps data from one
regular language to another via a user-defined
function.  For instance,
if \codefont{comma_name} describes the format for ``Conway, J. H.''
and \codefont{space_name} describes the format for ``J. H. Conway''
and \codefont{s_to_c} is a function from one to the other then
the following instance of squash creates the desired canonizer.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma_name} and \codefont{space_name} using the listed example
transformation as a guide.  The second line simply
extracts the left-to-right transformation from the lens, which is what we need for squash.

@@

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
quotient lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
\item \kw{del} $s$: delete the constant string $s$ in the left-to-right direction; produce the constant string $s$ in the right-to-left direction.
\item \kw{ins} $s$: produce the string $s$ in the left-to-right direction;
delete the string $s$ in the right-to-left direction.
\item \kw{id} $r$: copy the text matching $r$ from left-to-right (also from
right-to-left) \dpw{this was not the exact syntax.  fix this.}
\end{itemize}
%
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

Intuitively, our QRE lenses are bijective lenses with ``canonizers at the
edges''---Figure~\ref{fig:attheedges} presents the architecture of each lens.
Every QRE lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the forward direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.

>>>>

\emph{Lenses} define computations that can be read in two directions,
the get direction (forwards) and the put direction (backwards).  The simplest lenses define bijections between source and
data, but we are interested in \emph{quotient lenses}, which define
bijections modulo some equivalence relation.  The remains of this section
describe and motivate a new language and system for defining such lenses.

@@

While a projection chooses a single canonical representative for all elements
of a regular language, the \kw{squash} combinator maps data from one
regular language to another via a user-defined
function.  For instance,
if \codefont{comma\_name} describes the format for ``Conway, J. H.''
and \codefont{space\_name} describes the format for ``J. H. Conway''
and \codefont{s\_to\_c} is a function from one to the other then
the following instance of squash creates the desired canonizer.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma_name} and \codefont{space_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for squash.

@@

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
quotient lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
\item \kw{del} $s$: delete the constant string $s$ in the get direction;
produce the constant string $s$ in the put direction.
\item \kw{ins} $s$: produce the string $s$ in the get direction;
delete the string $s$ in the put direction.
\item \kw{id} $r$: copy the text matching $r$ in both directions \dpw{this was not the exact syntax.  fix this.}
\end{itemize}
%
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

Intuitively, our QRE lenses are bijective lenses with ``canonizers at the
edges''---Figure~\ref{fig:attheedges} presents the architecture of each lens.
Every QRE lens $q$ has a type $q: c \Leftrightarrow c'$ where $c$ and $c'$ are QREs.
In the get direction, a QRE lens $q: c \Leftrightarrow c'$ uses the source
QRE $c$ to compute a unique representative for the data modulo the equivalence
relation defined by $c$ and then applies the $\get$ function of a bijective
lens $\ell$ to this representative. In the backward direction, $q$ operates
similarly,  but using the view QRE $c'$ and the $\lput$ function of $\ell$.
